---
title: "Modelling responses to a stochastic Ebola virus epidemic"
output:
  bookdown::html_vignette2:
    fig_caption: yes
    code_folding: show
pkgdown:
  as_is: true
bibliography: references.json
link-citations: true
vignette: >
  %\VignetteIndexEntry{Modelling responses to a stochastic Ebola virus epidemic}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 5,
  fig.height = 4,
  dpi = 300
)
```

::: {.alert .alert-warning}
**New to _epidemics_?** It may help to read the ["Get started"](epidemics.html) vignette first!
:::

This vignette shows how to model an epidemic in which **stochasticity is expected to play an important role.**
This is often the case with outbreaks of infections such as Ebola virus disease (EVD, or simply, ebola).

_epidemics_ includes a discrete time, stochastic compartmental model of ebola based on a consensus model presented in @li2019, with six compartments --- susceptible, exposed, infectious, hospitalised, funeral, and removed.
The transitions between compartments are based on an Erlang model developed by @getz2018 for the 2014 West African EVD outbreak, with the code adapted from [Epirecipes](http://epirecip.es/epicookbook/chapters/erlang/r).

The model does not include demographic variation in contacts, as ebola spreads primarily among contacts caring for infected individuals, making age or demographic structure less important.

This model can currently accommodate interventions on model rates only (as there are no demographic groups for contacts interventions).

```{r setup}
# some initial setup to load necessary packages
library(epidemics)
library(dplyr)
library(ggplot2)
```

## Prepare population and initial conditions

Prepare population data --- for this model, we only need the total population size and the initial conditions.

We can create a `<population>` object with dummy values for the contact matrix, and assign the total size to the demography vector.

We prepare a population with a total size of 14 million, corresponding to the population of Guinea, a West African country where the 2014 Ebola virus epidemic originated.

We assume that 1 person is initially infected and infectious, and 10 other people have been exposed and is yet to become infectious.

```{r}
population_size <- 14e6

# prepare initial conditions as proportions
initial_conditions <- c(
  S = population_size - 11, E = 10, I = 1, H = 0, F = 0, R = 0
) / population_size
```


```{r}
# prepare a <population> object
guinea_population <- population(
  name = "Guinea",
  contact_matrix = matrix(1), # note dummy value
  demography_vector = 14e6, # 14 million, no age groups
  initial_conditions = matrix(
    initial_conditions,
    nrow = 1
  )
)

guinea_population
```

## Prepare model parameters

We use the default model parameters, beginning with an $R_0$ taken from the value estimated for ebola in Guinea [@althaus2014].
We assume that ebola has a mean infectious period of 12 days, and that the time between exposure and symptom onset --- the pre-infectious period --- is 5 days.

Together, these give the following values:

 - Transmissibility ($\beta$): 0.125,
 
 - Infectiousness rate ($\gamma^E$ in @getz2018): 0.4,
 
 - Removal rate ($\gamma^I$ in @getz2018): 0.1667

This model does not yet have an explicit "deaths" compartment, but rather, the "removed" compartment holds all individuals who have recovered and who have died and been buried safely.
Functionally, they are similar as they are not part of the model dynamics.

We assume that only 10% of infectious individuals are hospitalised, and that hospitalisation achieves a modest 30% reduction in transmission between hospitalised individuals and those still susceptible.

We assume also that any deaths in hospital lead to ebola-safe funerals, such that no further infections result from them.

We assume that infectious individuals who are not hospitalised die in the community, and that their funerals are potentially not ebola-safe.
We assume that the transmission rate of ebola at funerals is 50% of the baseline transmission rate.
This can also be interpreted as the proportion of funerals that are ebola-safe.

## Run epidemic model

We run the model using the function `model_ebola_r()`.
While there is a C++ equivalent of this model, it was not found to be much faster than the R-only version, and is not made available to users.

The model is run for 100 days, with data reported on each day.
This is an appropriate period over which to obtain initial predictions for the outbreak, and after which response efforts such as non-pharmaceutical interventions and vaccination campaigns are likely to be launched to bring the outbreak under control.

```{r}
# set seed to get consistent results
set.seed(0)

# run the epidemic model using `epidemic`
# we call the model "ebola" from the library
output <- model_ebola_r(
  population = guinea_population,
  time_end = 100
)

# View output
output

# get model data
data <- get_parameter(output, "data")
```

## Prepare data and visualise infections

We plot the ebola epidemic over time, showing the number of individuals exposed and infectious.

```{r class.source = 'fold-hide', fig.cap="Model results from a single run showing the number of individuals in the exposed and infectious compartments over 100 days of the outbreak. The model assumes that 1 initially infectious person has exposed 10 others."}
# plot figure of epidemic curve
filter(data, compartment %in% c("exposed", "infectious")) %>%
  ggplot(
    aes(
      x = time,
      y = value,
      colour = compartment
    )
  ) +
  geom_line() +
  scale_y_continuous(
    labels = scales::comma
  ) +
  scale_colour_brewer(
    palette = "Dark2",
    name = NULL,
    labels = c("Exposed", "Infectious")
  ) +
  expand_limits(
    x = c(0, 101)
  ) +
  theme_bw() +
  theme(
    legend.position = "top"
  ) +
  labs(
    x = "Simulation time (days)",
    linetype = "Compartment",
    y = "Individuals"
  )
```

## Consensus estimates from multiple model iterations

Since the ebola model is a stochastic one, model results will vary in each run.
We can run the model multiple times --- here, 100 times --- and plot the number of infectious people, to get a consensus idea of what an ebola outbreak might look like.

```{r}
# set seed for consistent output
set.seed(1)

# run the simulation 100 times
output <- lapply(
  X = seq(100),
  FUN = function(x) {
    data <- model_ebola_r(
      population = guinea_population,
      time_end = 100
    )
  }
)

# access the data from the model ojects and create a single data.frame
data_timeseries <- Map(output, seq(100), f = function(out, i) {
  data <- get_parameter(out, "data")

  # add replicate number and return data
  data$replicate <- i
  data
}) %>%
  bind_rows()
```

We plot the data.

```{r class.source = 'fold-hide', fig.cap="Model results from 100 runs showing the number of individuals in the infectious compartments over 100 days of each outbreak. All models assume that 1 initially infectious person has exposed 10 others. A simple summary using the mean (black line with standard error as shaded red regions), plotted over the individual model runs (grey) show that there can be wide variation in individual outbreak trajectories."}
filter(data_timeseries, compartment == "infectious") %>%
  ggplot() +
  geom_line(
    aes(time, value, group = replicate),
    alpha = 0.3, colour = "grey50"
  ) +
  stat_summary(
    aes(time, value),
    geom = "ribbon",
    fill = "red", alpha = 0.5
  ) +
  stat_summary(
    aes(time, value),
    geom = "line"
  ) +
  scale_y_continuous(
    labels = scales::comma
  ) +
  expand_limits(
    x = c(0, 101)
  ) +
  coord_cartesian(
    expand = FALSE
  ) +
  theme_bw() +
  theme(
    legend.position = "top"
  ) +
  labs(
    x = "Simulation time (days)",
    y = "Individuals infectious with ebola"
  )
```

We observe that model stochasticity leads to wide variation in model outcomes, and not all simulation replicates lead to exponential growth of the outbreak within the first 100 days.
However, in a number of replicates, the outbreak has only a low daily incidence within the simulation time, although these replicates might show exponential growth later.

We can find the 'final size' of each replicate using the `epidemic_size()` function on the original simulation data.
Note that the final size here refers to the final size after 365 days, which is the duration of our simulation.

```{r}
# apply the function over each replicate
data_final_size <- lapply(
  X = output,
  FUN = epidemic_size, include_deaths = FALSE, by_group = FALSE
)
data_final_size <- unlist(data_final_size)
```

```{r class.source = 'fold-hide', fig.cap="Histogram of the final outbreak sizes of 100 ebola outbreaks. These are taken from the 100 model runs presented in the earlier figure, and show that most outbreaks may affect fewer than 500 individuals within their first 100 days."}
# plot a histogram of the replicates
ggplot() +
  geom_histogram(
    aes(data_final_size),
    fill = "steelblue"
  ) +
  scale_x_continuous(
    labels = scales::comma
  ) +
  coord_cartesian(
    expand = FALSE
  ) +
  theme_bw() +
  theme(
    legend.position = "top"
  ) +
  labs(
    x = "Epidemic size at 100 days",
    y = "Number of replicates"
  )
```

We see that while some replicates affect over 600 individuals within the first 100 days, most replicates have fewer than 500 cases, suggesting that responses to halt the spread of ebola could be quite effective in most real outbreaks.

## Applying interventions that reduce transmission

An intervention that reduces transmission can be applied by passing a `<rate_intervention>` on the transmission rate parameter, `beta`.

We begin with an initial simple example that models an intervention that begins 15 days into the outbreak, and reduces transmission by 20%.

```{r}
# create an intervention on the transmission rate
reduce_transmission <- intervention(
  type = "rate",
  time_begin = 15, time_end = 100, reduction = 0.2
)
```

We can pass the intervention to the model function's `intervention` argument.

```{r}
# set a seed for comparison with the baseline model
set.seed(0)

# run the epidemic model and save data as the baseline
data_baseline <- model_ebola_r(
  population = guinea_population,
  time_end = 100
) %>%
  get_parameter("data")

set.seed(0)
# note that the intervention is passed within a list,
# where it is named for the rate it is targeting
data_intervention <- model_ebola_r(
  population = guinea_population,
  intervention = list(
    transmissibility = reduce_transmission
  ),
  time_end = 100
) %>%
  get_parameter("data")

# assign scenario names
data_baseline$scenario <- "baseline"
data_intervention$scenario <- "intervention"

# bind the data together
data_scenarios <- bind_rows(data_baseline, data_intervention)
```

```{r class.source = 'fold-hide', fig.cap="Effect of implementing an intervention that reduces transmission by 30% during an ebola outbreak. The intervention begins on the 15th day (dotted vertical line), and is active for the remainder of the model duration. Applying this intervention leads to many fewer individuals infectious with ebola over the outbreak."}
filter(data_scenarios, compartment == "removed") %>%
  ggplot() +
  geom_vline(
    xintercept = 15,
    linetype = "dotted"
  ) +
  geom_line(
    aes(time, value, colour = scenario)
  ) +
  scale_colour_brewer(
    palette = "Set1",
    name = "Scenario",
    labels = c("Baseline", "Intervention")
  ) +
  theme_bw() +
  theme(
    legend.position = "top"
  ) +
  labs(
    x = "Days",
    y = "Outbreak size"
  )
```

We see that applying an intervention that reduces transmission can be effective in reducing the number of individuals infected over the outbreak.

## Modelling the roll-out of vaccination

We can also model the rollout of vaccination against EVD, but because this model is structured differently from other models in _epidemics_, vaccination must be modelled differently too.

`model_ebola_r()` does not accept a vaccination regime as a `<vaccination>` object, but we can still model the effect of vaccination as a gradual decrease in the transmission rate $\beta$ over time.

This is done by using the _time dependence_ functionality of _epidemics_.

::: {.alert .alert-warning}
**New to implementing parameter time dependence in _epidemics_?** It may help to read the [vignette on time dependence functionality](time_dependence.html) first!
:::

We first define a function suitable for the `time_dependence` argument.
This function assumes that the baseline transmission rate of ebola decreases over time, with a 5% reduction each day due to vaccination.

```{r}
# we assume a maximum time of 100 days, and a 5% daily reduction
# we assume that this vaccination begins on the 15th day
time_dep_vax <- function(
    time, x, max_time = 100, time_start = 15,
    reduction = 0.05) {
  if (time < time_start) {
    x
  } else {
    x * ((1.0 - reduction)^(time / max_time))
  }
}
```

```{r}
# set a seed for comparison with the baseline model
set.seed(0)
# note that the intervention is passed within a list,
# where it is named for the rate it is targeting
data_apply_vax <- model_ebola_r(
  population = guinea_population,
  time_dependence = list(
    transmissibility = time_dep_vax
  ),
  time_end = 100
) %>%
  get_parameter("data")

# assign scenario name
data_apply_vax$scenario <- "apply_vaccination"

# bind the data together
data_scenarios <- bind_rows(data_baseline, data_apply_vax)
```

```{r class.source = 'fold-hide', fig.cap="Effect of implementing a vaccination regime that gradually reduces ebola transmission over time, using the time dependence functionality."}
filter(data_scenarios, compartment == "removed") %>%
  ggplot() +
  geom_vline(
    xintercept = 15,
    linetype = "dotted"
  ) +
  geom_line(
    aes(time, value, colour = scenario)
  ) +
  scale_colour_brewer(
    palette = "Set1",
    name = "Scenario",
    labels = c("Baseline", "Vaccination campaign")
  ) +
  theme_bw() +
  theme(
    legend.position = "top"
  ) +
  labs(
    x = "Days",
    y = "Outbreak size"
  )
```

Similar functionality can be used to model other parameters more flexibly than allowed by the `<rate_intervention>` class.

## Modelling a multi-pronged ebola response

Since EVD is a severe disease with a high case fatality risk, responses to an outbreak may require the simultaneous implementation of a number of measures to reduce transmission to end the outbreak quickly.

We can use the EVD model and existing functionality in _epidemics_ to model the implementation of a multi-pronged response to ebola that aims to improve:

- detection and treatment of cases,

- safety of hospitals and the risk of community transmission, and

- safety of funeral practices to reduce the risk of transmission from dead bodies.

Here, we show the effects of these interventions separately, before showing their combined effect.

### Improving hospitalisation

We first model an intervention that begins 15 days into the outbreak, and reduces the proportion of individuals who are infectious in the community by 30%, thus improving the proportion that are hospitalised.

```{r}
# create an intervention on the transmission rate
improve_hospitalisation <- intervention(
  type = "rate",
  time_begin = 15, time_end = 100, reduction = 0.3
)
```

We can pass the interventions to the model function's `intervention` argument as a named list, with the names indicating the model parameters to target.

```{r}
# set a seed for comparison with the baseline model
set.seed(0)
# note that the intervention is passed within a list,
# where it is named for the rate it is targeting
data_improve_hosp <- model_ebola_r(
  population = guinea_population,
  intervention = list(
    prop_community = improve_hospitalisation
  ),
  time_end = 100
) %>%
  get_parameter("data")

# assign scenario name
data_improve_hosp$scenario <- "improve_hosp"

# bind the data together
data_scenarios <- bind_rows(data_baseline, data_improve_hosp)
```

```{r class.source = 'fold-hide', fig.cap="Effect of implementing an intervention that reduces the proportion of infectious cases in the community by transferring them to a hospitalisation setting."}
filter(data_scenarios, compartment == "removed") %>%
  ggplot() +
  geom_vline(
    xintercept = 15,
    linetype = "dotted"
  ) +
  geom_line(
    aes(time, value, colour = scenario)
  ) +
  scale_colour_brewer(
    palette = "Set1",
    name = "Scenario",
    labels = c("Baseline", "Improved hospitalisation")
  ) +
  theme_bw() +
  theme(
    legend.position = "top"
  ) +
  labs(
    x = "Days",
    y = "Outbreak size"
  )
```

### Improve ETU safety

We next model an intervention to improve the safety of ETUs by reducing transmission risk by 70%, alongside the earlier improvement in hospitalisation.

```{r}
# create an intervention on the transmission rate
improve_etu_safety <- intervention(
  type = "rate",
  time_begin = 15, time_end = 100, reduction = 0.7
)
```

```{r}
# set a seed for comparison with the baseline model
set.seed(0)
# note that the intervention is passed within a list,
# where it is named for the rate it is targeting
data_improve_etu <- model_ebola_r(
  population = guinea_population,
  intervention = list(
    etu_risk = improve_etu_safety
  ),
  time_end = 100
) %>%
  get_parameter("data")

# assign scenario name
data_improve_etu$scenario <- "improve_etu"

# bind the data together
data_scenarios <- bind_rows(data_baseline, data_improve_etu)
```

```{r class.source = 'fold-hide', fig.cap="Effect of implementing an intervention that reduces the risk of ebola transmission in a hospitalisation context."}
filter(data_scenarios, compartment == "removed") %>%
  ggplot() +
  geom_vline(
    xintercept = 15,
    linetype = "dotted"
  ) +
  geom_line(
    aes(time, value, colour = scenario)
  ) +
  scale_colour_brewer(
    palette = "Set1",
    name = "Scenario",
    labels = c("Baseline", "Improve ETU safety")
  ) +
  theme_bw() +
  theme(
    legend.position = "top"
  ) +
  labs(
    x = "Days",
    y = "Outbreak size"
  )
```

Note that the reason that this intervention has little to no effect is that it is only likely to be effective when the proportion of cases hospitalised is high.
In this model, the large majority of cases remain in the community, and improving the safety of ETUs is not an effective intervention.

### Improve funeral safety

We next model an intervention to improve the safety of funeral practices, by reducing transmission risk associated with funerals by 50%.

```{r}
# create an intervention on the transmission rate
reduce_funeral_risk <- intervention(
  type = "rate",
  time_begin = 15, time_end = 100, reduction = 0.5
)
```

```{r}
# set a seed for comparison with the baseline model
set.seed(0)
# note that the intervention is passed within a list,
# where it is named for the rate it is targeting
data_improve_funeral <- model_ebola_r(
  population = guinea_population,
  intervention = list(
    funeral_risk = reduce_funeral_risk
  ),
  time_end = 100
) %>%
  get_parameter("data")

# assign scenario name
data_improve_funeral$scenario <- "improve_funeral"

# bind the data together
data_scenarios <- bind_rows(data_baseline, data_improve_funeral)
```

```{r class.source = 'fold-hide', fig.cap="Effect of implementing an intervention that reduces the risk of ebola transmission in a funeral context."}
filter(data_scenarios, compartment == "removed") %>%
  ggplot() +
  geom_vline(
    xintercept = 15,
    linetype = "dotted"
  ) +
  geom_line(
    aes(time, value, colour = scenario)
  ) +
  scale_colour_brewer(
    palette = "Set1",
    name = "Scenario",
    labels = c("Baseline", "Improve funeral safety")
  ) +
  scale_y_sqrt() +
  theme_bw() +
  theme(
    legend.position = "top"
  ) +
  labs(
    x = "Days",
    y = "Outbreak size"
  )
```

### Combined interventions

```{r}
# set a seed for comparison with the baseline model
set.seed(0)
# note that the intervention is passed within a list,
# where it is named for the rate it is targeting
data_combined_interventions <- model_ebola_r(
  population = guinea_population,
  intervention = list(
    transmissibility = reduce_transmission,
    prop_community = improve_hospitalisation,
    etu_risk = improve_etu_safety,
    funeral_risk = reduce_funeral_risk
  ),
  time_end = 100
) %>%
  get_parameter("data")

# assign scenario name
data_combined_interventions$scenario <- "combined_interventions"

# bind the data together
data_scenarios <- bind_rows(data_baseline, data_combined_interventions)
```

```{r class.source = 'fold-hide', fig.cap="Effect of implementing multiple simultaneous interventions to reduce transmission during an ebola outbreak, all beginning on the 15th day (dotted vertical line), and remaining active for the remainder of the model duration. Applying these interventions substantially reduces the final size of the outbreak, with a potential plateau in the outbreak size reached at 100 days."}
filter(data_scenarios, compartment == "removed") %>%
  ggplot() +
  geom_vline(
    xintercept = 15,
    linetype = "dotted"
  ) +
  geom_line(
    aes(time, value, colour = scenario)
  ) +
  scale_colour_brewer(
    palette = "Set1",
    name = "Scenario",
    labels = c("Baseline", "Combined interventions")
  ) +
  theme_bw() +
  theme(
    legend.position = "top"
  ) +
  labs(
    x = "Days",
    y = "Outbreak size"
  )
```

## Details: Discrete-time Ebola virus disease model

This model has compartments adopted from the consensus model for Ebola virus disease presented in Li et al. (2019), and with transitions between epidemiological compartments modelled using Erlang sub-compartments adapted from Getz and Dougherty (2018); see **References**.

The R code for this model is adapted from code by Ha Minh Lam and initially made available on _Epirecipes_ (https://github.com/epirecipes/epicookbook) under the MIT license.

The transition rates between the exposed and infectious, and infectious and funeral compartments (and also hospitalised to removed), $\gamma^E$ and $\gamma^I$ in Getz and Dougherty's notation, are passed by the user as the `infectiousness_rate` and `removal_rate` respectively.

The shape of the Erlang distributions of passage times through the exposed and infectious compartments ($k^E$ and $k^I$) are recommended to be set to 2 as a sensible choice, which is the default value for the `erlang_sbubcompartments` argument, but can be allowed to vary (but not independently).

Getz and Dougherty's equation (6) gives the relationship between these parameters and the mean pre-infectious $\rho^E$ and infectious $\rho^I$ periods.

$$\gamma^E = \dfrac{k^E}{\rho^E} = \dfrac{2}{\rho^E} ~\text{and}~
\gamma^I = \dfrac{k^I}{\rho^I} = \dfrac{2}{\rho^I}$$

In this discrete time model, $\gamma^E$ and $\gamma^I$ are used to determine the passage times of newly exposed or infectious individuals through their respective compartments (thus allowing for variation in passage times).

### Hospitalisation, funerals, and removal

@li2019 present a consensus model for EVD in which individuals can follow multiple pathways from the infectious compartment --- to hospitalisation, funeral transmission, and safe burial or recovery (removed), with the possibility of skipping some compartments in this sequence (e.g. directly from infectious to removed, infectious to hospitalisation to removed etc.).
This model simplifies these transitions to only two pathways:

1. Infectious → funeral → removed (safe burial)

2. Infectious → hospitalised → removed (recovery or safe burial)

A proportion, `1.0 - prop_community`, of infectious individuals are transferred to the hospitalised compartment in each timestep, This compartment represents Ebola Treatment Units (ETUs), and individuals in the hospitalised compartment are considered to be infectious but no longer in the community.

The passage time of individuals in the hospitalised compartment is similar to that of individuals in the infectious compartment (i.e., infectious in the community), which means that an infectious individual with $N$ timesteps before exiting the infectious compartment will exit the hospitalised compartment in the same time.

Hospitalised individuals can contribute to transmission of Ebola to susceptibles depending on the value of $p_\text{ETU}$, which is the probability (or proportion) of hospitalised cases contributing to the infection of susceptibles.
This is interpreted as the relative risk of Ebola virus treatment units (ETUs), with a range of 0.0 -- 1.0, where 0.0 indicates that hospitalisation prevents onward transmission entirely, while 1.0 indicates that hospitalisation does not reduce transmission at all.
This is passed as the argument `etu_risk`.

We assume that deaths in hospital lead to Ebola-safe funerals, and individuals exiting the hospitalised compartment move to the 'removed' compartment, which holds both recoveries and deaths.

We assume that infectious individuals who are not hospitalised die, and that deaths outside of hospital lead to funerals that are potentially not Ebola-safe, and the contribution of funerals to Ebola transmission is determined by $p_\text{funeral}$, which can be interpreted as the relative risk of funerals, and is passed as the `funeral_risk` argument.
Values closer to 0.0 indicate that there is low to no risk of Ebola transmission at funerals, while values closer to 1.0 indicate that the risk is similar to that of transmission in the community.

Individuals are assumed to spend only a single timestep in the funeral transmission compartment, before they move into the 'removed' compartment. Individuals in the 'removed' compartment do not affect model dynamics.

## References
