---
title: "Modelling the effect of a vaccination campaign"
output:
  bookdown::html_vignette2:
    fig_caption: yes
    code_folding: show
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{Modelling the effect of a vaccination campaign}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi = 150
)
```

```{r setup}
library(epidemics)
```

## Prepare population and initial conditions

Prepare population and contact data.

```{r}
# load contact and population data from socialmixr::polymod
polymod <- socialmixr::polymod
contact_data <- socialmixr::contact_matrix(
  polymod,
  countries = "United Kingdom",
  age.limits = c(0, 20, 65),
  symmetric = TRUE
)

# prepare contact matrix
contact_matrix <- t(contact_data$matrix)

# prepare the demography vector
demography_vector <- contact_data$demography$population
names(demography_vector) <- rownames(contact_matrix)
```

Prepare initial conditions for each age group.

```{r}
# initial conditions
initial_i <- 1e-6
initial_conditions <- c(
  S = 1 - initial_i, E = 0, I = initial_i, R = 0, V = 0
)

# build for all age groups
initial_conditions <- rbind(
  initial_conditions,
  initial_conditions,
  initial_conditions
)

# assign rownames for clarity
rownames(initial_conditions) <- rownames(contact_matrix)
```

Prepare a population as a `population` class object.

```{r}
uk_population <- population(
  name = "UK",
  contact_matrix = contact_matrix,
  demography_vector = demography_vector,
  initial_conditions = initial_conditions
)
```

## Prepare a vaccination campaign

Prepare a vaccination campaign targeting individuals aged over 65.

```{r}
# prepare a vaccination object
vaccinate_elders <- vaccination(
  name = "vaccinate elders",
  time_begin = matrix(100, nrow(contact_matrix)),
  time_end = matrix(250, nrow(contact_matrix)),
  nu = matrix(c(0.0001, 0, 0))
)

# view vaccination object
vaccinate_elders
```

::: {.alert .alert-warning}
**Note that** when the vaccination rate is high, the number of individuals who could _potentially_ transition from 'susceptible' to 'vaccinated' may be greater than the number of individuals in the 'susceptible' compartment: $dS_{i_{t}} = \nu_{i_t}S_{i_{0}} > S_{i_{t}}$ for each demographic group $i$ at time $t$.

This could realistically occur when there are more doses available than individuals eligible to receive them, for instance towards the end of an epidemic.

_epidemics_ automatically handles such situations by setting $dS_{i_{t}}$ to be the minimum of doses or eligible individuals: : $dS_{i_t} = \text{Min}(\nu_{i_t}S_{i_0}, S_{i_t})$, such that $S_{i_{t+1}}$ does not take a negative value.
:::

The `<vaccination>` object can be passed to the `vaccination` argument of a model function call.

```{r}
# run an epidemic model using `epidemic`
output <- model_default(
  population = uk_population,
  vaccination = vaccinate_elders,
  time_end = 600, increment = 1.0
)
```

::: {.alert .alert-info}
**Note that** vaccination modelling is currently only supported for the 'default' (`model_default()`) and 'Vacamole' (`model_vacamole()`) models.
:::
