---
title: "Modelling a vaccination campaign"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{modelling_vaccination}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(epidemics)
library(data.table)
library(ggplot2)
library(stringr)
```

## Prepare population and initial conditions

Prepare population and contact data.

```{r}
# load contact and population data from socialmixr::polymod
polymod <- socialmixr::polymod
contact_data <- socialmixr::contact_matrix(
  polymod,
  countries = "United Kingdom",
  age.limits = c(0, 20, 65),
  symmetric = TRUE
)

# prepare contact matrix
contact_matrix <- t(contact_data$matrix)

# prepare the demography vector
demography_vector <- contact_data$demography$population
names(demography_vector) <- rownames(contact_matrix)
```

Prepare initial conditions for each age group.

```{r}
# initial conditions
initial_i <- 1e-6
initial_conditions <- c(
  S = 1 - initial_i, E = 0, I = initial_i, R = 0, V = 0
)

# build for all age groups
initial_conditions <- rbind(
  initial_conditions,
  initial_conditions,
  initial_conditions
)

# assign rownames for clarity
rownames(initial_conditions) <- rownames(contact_matrix)
```

Prepare a population as a `population` class object.

```{r}
uk_population <- population(
  name = "UK",
  contact_matrix = contact_matrix,
  demography_vector = demography_vector,
  initial_conditions = initial_conditions
)
```

## Prepare pathogen parameters

```{r}
# Prepare epidemiological parameters
r0 <- rep(1.5, nrow(contact_matrix))
preinfectious_period <- rep(3, nrow(contact_matrix))
infectious_period <- rep(7, nrow(contact_matrix))
```

## Prepare a vaccination campaign

Prepare a vaccination campaign targeting individuals aged over 65.

```{r}
# prepare a vaccination object
vaccinate_elders <- vaccination(
  name = "vaccinate elders",
  time_begin = rep(100, nrow(contact_matrix)),
  time_end = rep(250, nrow(contact_matrix)),
  nu = c(0.0001, 0.0001, 0.0001)
)

# view vaccination object
vaccinate_elders
```

## Run epidemic model

```{r}
# run an epidemic model using `epidemic_cpp`
output <- epidemic_cpp(
  population = uk_population,
  r0 = r0,
  preinfectious_period = preinfectious_period,
  infectious_period = infectious_period,
  vaccination = vaccinate_elders,
  time_end = 600, increment = 1.0
)
```

## Prepare data and visualise infections

```{r}
# make long format data
data <- melt(
  output[, .SD, .SDcols = names(output) %like%
    "infectious|exposed|vaccinated|time"],
  id.vars = "time"
)

# extract age group and compartment
data[, c("compartment", "age_group") := list(
  str_extract(variable, "(\\w+)(?=_)"),
  str_extract(variable, "\\d{1}")
)]

# assign age groups
age_groups <- data.table(
  age_group = as.character(seq_along(demography_vector)),
  age_group_name = names(demography_vector)
)

data <- merge(data, age_groups)
```

Plot epidemic over time, showing only the number of individuals in the exposed, infected, and vaccinated compartments.

```{r}
# plot figure of epidemic curve
ggplot(
  data,
  aes(
    x = time,
    y = value,
    col = age_group_name,
    linetype = compartment
  )
) +
  geom_line() +
  scale_y_continuous(
    labels = scales::comma
  ) +
  scale_colour_brewer(
    palette = "Dark2",
    name = "Age group"
  ) +
  expand_limits(
    y = c(0, 500e3)
  ) +
  coord_cartesian(
    expand = FALSE
  ) +
  theme_classic() +
  theme(
    legend.position = "top"
  ) +
  labs(
    x = "Simulation time (days)",
    linetype = "Compartment",
    y = "Individuals"
  )
```
