---
title: "Modelling parameter uncertainty and epidemic scenarios"
output:
  bookdown::html_vignette2:
    fig_caption: yes
    code_folding: show
pkgdown:
  as_is: true
bibliography: references.json
link-citations: true
vignette: >
  %\VignetteIndexEntry{Modelling parameter uncertainty and epidemic scenarios}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4,
  dpi = 150
)
```

::: {.alert .alert-warning}
**New to _epidemics_?** It may help to read the ["Get started"](epidemics.html) vignette first!
:::

This vignette shows how _epidemics_ can conveniently be used to:

1. Include parameter uncertainty in an epidemic model;
2. Run multiple scenarios of epidemic response measures;
3. Include parameter uncertainty when running multiple scenarios.

::: {.alert .alert-warning}
**NOTE:** This vignette applies only to the deterministic ODE models --- the default, Vacamole, and diphtheria models.
The Ebola virus disease model included in the package is expected to receive this functionality in the near future.
:::

```{r setup}
# epi modelling
library(epidemics)
library(EpiEstim)

# data manipulation packages
library(dplyr)
library(tidyr)
library(ggplot2)

# for reproducibility
library(withr)
```

## Modelling parameter uncertainty

Uncertainty in the characteristics of an epidemic is a key element and roadblock in epidemic response [@shea2020]. Epidemic dynamics can be influenced by two main sources of uncertainty: intrinsic randomness in the transmission process, and uncertain knowledge of the parameters underlying the transmission process. Here we focus on the latter, i.e. uncertainty in the input parameters of an epidemic model.

_epidemics_ model functions can accept numeric vectors for all infection parameters; the model is run for each parameter combination using the same population and other model components (interventions, vaccinations) etc.

The code to set up a population is hidden here, but is similar to that found in the ["Getting started with epidemic scenario modelling components"](epidemics.html) vignette, and models the U.K. population with three age groups, and social contacts stratified by age.

```{r class.source = 'fold-hide'}
# load contact and population data from socialmixr::polymod
polymod <- socialmixr::polymod
contact_data <- socialmixr::contact_matrix(
  polymod,
  countries = "United Kingdom",
  age.limits = c(0, 20, 40),
  symmetric = TRUE
)

# prepare contact matrix
contact_matrix <- t(contact_data$matrix)

# prepare the demography vector
demography_vector <- contact_data$demography$population
names(demography_vector) <- rownames(contact_matrix)

# initial conditions
initial_i <- 1e-6
initial_conditions <- c(
  S = 1 - initial_i, E = 0, I = initial_i, R = 0, V = 0
)

# build for all age groups
initial_conditions <- rbind(
  initial_conditions,
  initial_conditions,
  initial_conditions
)
# assign rownames for clarity
rownames(initial_conditions) <- rownames(contact_matrix)
```

```{r}
# UK population created from hidden code
uk_population <- population(
  name = "UK",
  contact_matrix = contact_matrix,
  demography_vector = demography_vector,
  initial_conditions = initial_conditions
)
```

### Obtaining and passing a vector of infection parameters

For this example, we consider influenza with pandemic potential [@ghani2010], and prepare multiple samples of the estimated $R$.

We obtain the probability distribution function (PDF) from the distribution of the serial intervals; this is a Gamma distribution with shape $k$ = 2.622 and scale $\theta$ = 0.957 [@ghani2010].

::: {.alert .alert-info}
The [forthcoming Epiverse package _epiparameter_](https://epiverse-trace.github.io/epiparameter/) is expected to make it substantially easier to access and use epidemiological parameters, such as the serial interval, reported in the literature, making it easier to model scenarios differing in the intrinsic characteristics of the pathogen causing the outbreak.
:::

We use this PDF to estimate the $R$ of the 2009 influenza pandemic in the U.K., using [the _EpiEstim_ package](https://cran.r-project.org/package=EpiEstim).
We use the $R$ estimate (the mean and standard deviation) from _EpiEstim_ to generate 100 samples of $R$, assuming that it is normally distributed.

```{r fig.width=6, fig.height=3}
# Get 2009 influenza data for school in Pennsylvania
data(Flu2009)
flu_early_data <- filter(Flu2009$incidence, dates < "2009-05-10")

# get the PDF of the distribution of serial intervals
serial_pdf <- dgamma(seq(0, 25), shape = 2.622, scale = 0.957)
serial_pdf <- serial_pdf / sum(serial_pdf) # check sum to 1

# Use EpiEstim to estimate R with uncertainty
# Uses Gamma distribution by default
output_R <- estimate_R(
  incid = flu_early_data,
  method = "non_parametric_si",
  config = make_config(list(si_distr = serial_pdf))
)

# Plot output to visualise
plot(output_R, "R")
```

Generate 100 samples of $R$, and convert to the transmissibility parameter (often denoted $\beta$) by dividing by the infectious period of 7 days.

```{r}
# Generate 100 R samples
r_samples <- with_seed(
  seed = 1,
  rnorm(
    n = 100, mean = output_R$R$`Mean(R)`, sd = output_R$R$`Std(R)`
  )
)

infectious_period <- 7
beta <- r_samples / infectious_period
```

```{r}
# pass the vector of transmissibilities to the argument `transmissibility`
output <- model_default(
  population = uk_population,
  transmissibility = beta,
  time_end = 600
)

# view the output
head(output)
```

The output is a nested `<data.table>`, with the output of each run of the model for each unique `transmissibility` contained as a `<data.frame>` in the list column `"data"`.

::: {.alert .alert-info}
### Output type for vector parameter inputs

The output of `model_*()` when an infection parameter is passed as a vector is a nested `<data.table>`. This is similar to a nested `<tibble>`, and can be handled by popular data science packages, such as from the Tidyverse.

More on handling nested data can be found in this [section on list-columns in R for Data Science](https://r4ds.hadley.nz/rectangling.html#list-columns) and in the [documentation for nested data in the _tidyr_ package](https://tidyr.tidyverse.org/articles/nest.html). Equivalent operations are possible on `<data.table>`s directly; see [this R Bloggers post on unnesting data](https://www.r-bloggers.com/2019/10/much-faster-unnesting-with-data-table/).
:::

```{r}
# select the parameter set and data columns with dplyr::select()
# use tidyr to unnest the data column
data <- select(output, param_set, transmissibility, data) %>%
  unnest(data)
```

```{r class.source = 'fold-hide', fig.cap="Epidemic curves for the number of infectious individuals on each day of the epidemic; shaded region indicates uncertainty in the form of standard error."}
# plot the data
filter(data, compartment == "infectious") %>%
  ggplot() +
  stat_summary(
    geom = "ribbon",
    aes(time, value),
    fill = alpha("grey", 0.5)
  ) +
  stat_summary(
    geom = "line",
    aes(time, value)
  ) +
  scale_y_continuous(
    labels = scales::comma
  ) +
  labs(
    x = "Time", y = "Individuals infectious"
  ) +
  facet_wrap(~demography_group) +
  theme_classic()
```

::: {.alert .alert-info}
### Passing parameter sets

_epidemics_ model functions can accept multiple infection parameters as vectors, so long as one of the following rules is respected:

1. All infection parameter vectors must be of the same length;

  ```r
  beta <- rnorm(n = 100, mean, sd)
  gamma <- rnorm(n = 100, mean, sd) # the recovery rate
  
  model_default(
    population,
    transmissibility = beta,
    recovery_rate = gamma
  )
  ```

2. All but one infection parameter must be scalar values (single numbers). This is the worked-out example shown above.
:::

### Passing vectors of epidemic duration

_epidemics_ allows the duration of an model run to be varied, as this may be useful when examining how variation in the start time of an epidemic affects outcomes by a fixed end point.

```{r}
# draw samples of time_end
max_time <- 600
duration <- max_time - with_seed(seed = 1, {
  rnbinom(100, 1, 0.02)
})
```

```{r}
# pass the vector of durations to `time_end`
output <- model_default(
  population = uk_population,
  time_end = duration
)

# view the output
head(output)
```

::: {.alert .alert-warning}
**NOTE:** When the duration of the model runs is varied, each model output will have a potentially distinct number of rows.
:::

```{r message=FALSE, warning=FALSE}
# calculate the epidemic size to plot a distribution of sizes
epidemic_size_estimates <- select(output, param_set, data) %>%
  mutate(
    size = vapply(data, epidemic_size, by_group = FALSE, FUN.VALUE = 1)
  ) %>%
  select(-data)
```

```{r}
# view the range of epidemic sizes
range(epidemic_size_estimates$size)
```

## Modelling scenarios of epidemic response

Users may wish to model epidemic trajectories and outcomes under multiple scenarios of response measures.

_epidemics_ model functions can accept lists of intervention sets and vaccination regimes, allowing multiple intervention, vaccination, and combined intervention-and-vaccination sets to be modelled on the same population.

::: {.alert .alert-warning}
It may help to read the ["Modelling a non-pharmaceutical intervention targeting social contacts"](modelling_interventions.html) and the ["Modelling the effect of a vaccination campaign"](modelling_vaccination.html) vignettes first!
:::

::: {.alert .alert-info}
### Which model components can be passed as lists

_epidemics_ currently offers the ability to pass **intervention sets** and **vaccination regimes** as lists.

It is not possible to pass lists of _populations_, _seasonality_, or _population changes_ (for the diphtheria model).

In future, it may be possible to pass multiple populations as a list, to rapidly and conveniently model an epidemic on different populations, or to examine the effect of assumptions about the population's social contacts or demographic structure.
:::

### Creating a list of intervention sets

We shall create a list of 'intervention sets', each set representing a scenario of epidemic response measures.

**Note that** each intervention set is simply a list of `<intervention>` objects; typically, a single `<contacts_intervention>` and any `<rate_interventions>` on infection parameters.

```{r}
# create three distinct contact interventions
# prepare an intervention that models school closures for 180 days
close_schools <- intervention(
  name = "School closure",
  type = "contacts",
  time_begin = 60,
  time_end = 60 + 180,
  reduction = matrix(c(0.3, 0.01, 0.01))
)

# prepare an intervention which mostly affects adults 20 -- 65
intervention_duration <- 200
close_workplaces <- intervention(
  name = "Workplace closure",
  type = "contacts",
  time_begin = 80,
  time_end = 80 + intervention_duration,
  reduction = matrix(c(0.01, 0.3, 0.01))
)

# prepare a combined intervention
combined_intervention <- c(close_schools, close_workplaces)
```

```{r}
# create a mask-mandate rate intervention
# prepare an intervention that models mask mandates for 300 days
mask_mandate <- intervention(
  name = "mask mandate",
  type = "rate",
  time_begin = 60,
  time_end = 60 + 300,
  reduction = 0.1
)
```

Having prepared the interventions, we create intervention sets, and pass them to a model function.

```{r}
# create intervention sets, which are combinations of contacts and rate
# interventions
intervention_scenarios <- list(
  baseline = NULL,
  scenario_01 = list(
    contacts = close_schools
  ),
  scenario_02 = list(
    contacts = close_workplaces
  ),
  scenario_03 = list(
    contacts = combined_intervention
  ),
  scenario_04 = list(
    transmissibility = mask_mandate
  ),
  scenario_05 = list(
    contacts = close_schools,
    transmissibility = mask_mandate
  ),
  scenario_06 = list(
    contacts = close_workplaces,
    transmissibility = mask_mandate
  ),
  scenario_07 = list(
    contacts = combined_intervention,
    transmissibility = mask_mandate
  )
)
```

**Note that** there is no parameter uncertainty included here.

The output is a nested `<data.table>` as before.

```{r}
# pass the list of intervention sets to the model function
output <- model_default(
  uk_population,
  intervention = intervention_scenarios,
  time_end = 600
)

# examine the output
head(output)
```

::: {.alert .alert-info}
### Output type for list intervention inputs

The output of `model_*()` when either `intervention` or `vaccination` is passed a list, is a nested `<data.table>`.
This is similar to the output type when parameters are passed as vectors, but with the `"scenario"` column indicating each intervention set as a distinct scenario, which helps with grouping the model outputs in `"data"`.
:::

::: {.alert .alert-info}
### Combinations of intervention and vaccination scenarios

When either or both `intervention` and `vaccination` are passed lists (of intervention sets, or of vaccination regimes, respectively), the model is run for each combination of elements.
Thus for $M$ intervention sets and $N$ vaccination regimes, the model runs $M \times N$ times, and each combination is treated as a unique scenario.

**Note that** 'no intervention' and 'no vaccination' scenarios are not automatically created, and must be passed explicitly as `NULL` in the respective lists.

While the number of intervention and vaccination combinations are not expected to be very many, the addition of parameter uncertainty for each scenario (next section) may rapidly multiply the number of times the model is run.
**Users are advised** to be mindful of the number of scenario combinations they create.
:::

The unnesting of the output follows the same procedure as before, and is not shown here.

## Modelling epidemic response scenarios with parameter uncertainty

_epidemics_ allows parameter uncertainty to be combined with scenario models of multiple intervention sets or vaccination campaigns.

```{r}
# this example includes 100 samples of transmissibility for each intervention
# including the baseline
output <- model_default(
  uk_population,
  transmissibility = beta,
  intervention = intervention_scenarios,
  time_end = 600
)
```

The output is a nested `<data.table>` as before, and can be handled in the same way.

::: {.alert .alert-info}
### Output type for intervention and parameter set combinations

The output of `model_*()` when either `intervention` or `vaccination` is passed a list, and when any of the infection parameters is passed as a vector is a nested `<data.table>`.
The `"scenario"` column indicates each intervention set as a distinct scenario, while the `"param_set"` column indicates the parameter set used for the model run.

The **same parameters** are used for each scenario, allowing for comparability between scenarios at the aggregate level.

The output has $M \times N \times S$ rows (and nested model outputs), for $M$ intervention sets, $N$ vaccination regimes, and $S$ infection parameter sets.
:::

### Visualising outcomes across scenarios

Here, we show one way of visualising the epidemic size across scenarios, accounting for parameter uncertainty.

```{r message=FALSE}
# select the data, the parameter set, and the scenario
epidemic_size_estimates <- select(output, param_set, scenario, data) %>%
  mutate(
    size = vapply(data, epidemic_size, by_group = FALSE, FUN.VALUE = 1)
  ) %>%
  select(-data)
```

```{r class.source = 'fold-hide', fig.cap="The epidemic final size for multiple scenarios with parameter uncertainty in the transmissibility paramater."}
# set intervention labels
labels <- c(
  "No response", "Close schools", "Close workplaces", "Close both",
  "Mask mandate", "Masks + close schools", "Masks + close workplaces",
  "Masks + close both"
)

ggplot(epidemic_size_estimates) +
  geom_boxplot(
    aes(as.factor(scenario), size),
    fill = "grey"
  ) +
  scale_x_discrete(
    name = NULL,
    labels = labels,
    guide = guide_axis(n.dodge = 2)
  ) +
  scale_y_continuous(
    labels = scales::comma,
    name = "Epidemic final size"
  ) +
  theme_classic()
```
