---
title: "Modelling interventions on model rates"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modelling interventions on model rates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 5,
  fig.height = 4,
  dpi = 300
)
```

::: {.alert .alert-warning}
**New to _epidemics_, or to modelling non-pharmaceutical interventions?** It may help to read the ["Get started"](epidemics.html) and ["Modelling a non-pharmaceutical intervention"](modelling_interventions.html) vignettes first!
:::

```{r setup}
library(epidemics)
library(data.table)
library(ggplot2)
```

## Prepare population, initial conditions, and pathogen

We prepare population and contact data from the U.K., with epidemiological compartments matching the default epidemic model (SEIR-V).

We prepare an `infection` object to simulate a pathogen with an $R_0$ of 1.5 which is similar to pandemic influenza, an infectious period of 7 days, and a pre-infectious period of 3 days.

We assume that one in every million people has been infected and is infectious, translating to about 67 total infections for a U.K. population of 67 million.

The code for these steps is similar to that in the ["Getting started vignette"](epidemics.html) and is hidden here, although it can be expanded for reference.

```{r class.source = 'fold-hide'}
# load contact and population data from socialmixr::polymod
polymod <- socialmixr::polymod
contact_data <- socialmixr::contact_matrix(
  polymod,
  countries = "United Kingdom",
  age.limits = c(0, 20, 65),
  symmetric = TRUE
)

# prepare contact matrix
contact_matrix <- t(contact_data$matrix)

# prepare the demography vector
demography_vector <- contact_data$demography$population
names(demography_vector) <- rownames(contact_matrix)
```

```{r class.source = 'fold-hide'}
# initial conditions
initial_i <- 1e-4
initial_conditions <- c(
  S = 1 - initial_i, E = 0, I = initial_i, R = 0, V = 0
)

# build for all age groups
initial_conditions <- rbind(
  initial_conditions,
  initial_conditions,
  initial_conditions
)

# assign rownames for clarity
rownames(initial_conditions) <- rownames(contact_matrix)
```

```{r class.source = 'fold-hide'}
uk_population <- population(
  name = "UK",
  contact_matrix = contact_matrix,
  demography_vector = demography_vector,
  initial_conditions = initial_conditions
)
```

```{r class.source = 'fold-hide'}
# Prepare epidemiological parameters as an infection object
pandemic <- infection(
  r0 = 1.3,
  preinfectious_period = 1.5,
  infectious_period = 5
)
```

## Modelling an intervention on transmission rates

We model an intervention on the transmission rate $\beta$ that reduces it by 10%.
This could represent interventions such as requiring people to wear masks that reduce transmission.

```{r}
# prepare an intervention that models mask mandates for ~3 months (100 days)
mask_mandate <- intervention(
  name = "mask mandate",
  type = "rate",
  time_begin = 60,
  time_end = 60 + 100,
  reduction = 0.1
)

# examine the intervention object
mask_mandate

# check the object
is_intervention(mask_mandate)

is_contacts_intervention(mask_mandate)

is_rate_intervention(mask_mandate)
```

We simulate the epidemic for 200 days as we expect the intervention to spread disease incidence out over a longer period.

```{r}
# no intervention baseline scenario
data <- epidemic_default_cpp(
  population = uk_population,
  infection = pandemic,
  intervention = list(beta = mask_mandate),
  time_end = 200, increment = 1.0
)

# get new infections
data_infections <- new_infections(data, by_group = TRUE)
```

```{r}
ggplot(data_infections) +
  geom_line(
    aes(time, new_infections, col = demography_group)
  ) +
  coord_cartesian(
    expand = FALSE
  ) +
  annotate(
    geom = "rect",
    xmin = get_parameter(mask_mandate, "time_begin"),
    xmax = get_parameter(mask_mandate, "time_end"),
    ymin = 0, ymax = 100e3,
    fill = alpha("red", alpha = 0.2),
    lty = "dashed"
  ) +
  scale_y_continuous(
    labels = scales::comma
  ) +
  scale_colour_brewer(
    palette = "Dark2",
    name = "Age group"
  ) +
  expand_limits(
    y = c(0, 100e3)
  ) +
  coord_cartesian(
    expand = FALSE
  ) +
  theme_classic() +
  theme(
    legend.position = "top"
  ) +
  labs(
    x = "Simulation time (days)",
    linetype = "Compartment",
    y = "New infections"
  )
```

